# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

setting the directory:
```{r}
    currentDirectory <- getwd()
    newDirectory <- paste(currentDirectory,"/PeerAssessment1",sep="")
    if(!file.exists("PeerAssessment1")){dir.create("PeerAssessment1")}
    setwd(newDirectory)
```

downloading, unzipping and reading the data:
```{r}
    setInternet2()
    file <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(file,"data.zip")    
    unzip("data.zip")
    data <- read.csv("activity.csv")
```

loading required packages:
```{r}
    install.packages("lattice",repos="http://cran.us.r-project.org")
    library("lattice")
```

## What is mean total number of steps taken per day?

making copy of original data (for later computation of imputing missing values):
```{r}
    datazero <- data
```

excluding NA-entries (otherwise days with NA entries only get the value of "0" instead of "NA" in the following computation):
```{r}
    data<-data[complete.cases(data),]
```

computing distribution of total number of steps per day:
```{r}
    StepsPerDayTotal <- tapply(data$steps,data$date,sum,na.rm=T)
```

plotting the histogram:
```{r}
    hist(StepsPerDayTotal,xlab="Number of Steps",
    main="Distribution Total Number of Steps per Day")
```

computing and reporting mean **number** of **steps per day**:
```{r}
    StepsPerDayMean <- tapply(data$steps,data$date,mean,na.rm=T)
    StepsPerDayMean 
```


computing and reporting median **number** of **steps per day**:
```{r}
    StepsPerDayMedian <- tapply(data$steps,data$date,median,na.rm=T)
    StepsPerDayMedian
```

calculating **mean** number of **total steps over all days**:
```{r}
    MEAN <- mean(StepsPerDayTotal,na.rm=T)
    MEAN
```

calculating **median** number of **total steps over all days**:
```{r}
    MEDIAN <- median(StepsPerDayTotal,na.rm=T)
    MEDIAN
```

The **mean** and **median** number of total steps over all days are **`r MEAN`** and **`r MEDIAN`**.



## What is the average daily activity pattern?

calculating average number of steps taken per time interval, averaged across all days:
```{r}
    data <- datazero 
    IntervalAverages <- tapply(data$steps,data$interval,mean,na.rm=T)
```

plotting time series data for the average number of steps taken per time interval:
```{r}
    plot(x=names(IntervalAverages),y=IntervalAverages,type="l", xlab="Time Inteval", 
    ylab="Number of Steps", main="Number of Steps Averaged per Time Interval")
```

computing the maximum:
```{r}
    Maximum <- max(IntervalAverages)
    MaxPos <- IntervalAverages == Maximum
    (IntervalAverages[MaxPos])
    MaxInterval <- names(IntervalAverages[MaxPos])
```

The Interval **`r MaxInterval`** has on average the highest number of steps: **`r Maximum`**


## Imputing missing values

Creating data frame *dataNaRm* in which missing values are replaced with the related interval's average.

finding positions, intervals and the number of points for which no values were recorded:
```{r}
    MissingValues <- is.na(data$steps)
    NumberMissingValues <- sum(MissingValues)
    IntervalMissingValues <- data[MissingValues,"interval"]
    NumberMissingValues
```
For **`r NumberMissingValues`** of the intervals in the dataset no values were recorded.


picking the mean interval values for those points from the *IntervalAverages* calculated above:
```{r}
    MeanValues <- sapply(IntervalMissingValues,function(MVInt)
      {MVIntMean <- MVInt==names(IntervalAverages);IntervalAverages[MVIntMean]})
```

creating new data frame *dataNaRm* where those missing values are replaced by the related interval means:
```{r}
    dataNaRm <- data
    dataNaRm[MissingValues,"steps"] <- MeanValues
```

### Ploting the histogram with the *new* data

histogram with new (NA-imputed) data compared to histogram with old (NA-removed) data for comparison
```{r}
    ## new hist (NA-imputed)
    hist(tapply(dataNaRm$steps,dataNaRm$date,sum,na.rm=T),xlab="Number of Steps",
    main="Distribution Total Number of Steps per Day with Missing Values replaced",
    col = "blue",breaks = 10)
    
    ## old hist (NA-removed)
    data<-datazero[complete.cases(datazero),]
    hist(tapply(data$steps,data$date,sum,na.rm=T),xlab="Number of Steps",
    main="Distribution Total Number of Steps per Day with Missing Values replaced",add =     
      T, col = "lightblue",breaks = 10)
    
    ## adding legend
    legend("right",legend = c("change by NA-imputing")
    ,fill   = c("blue")
      )
```

computing and reporting the **new** mean number of **steps per day**:
```{r}
    StepsPerDayMean <- tapply(dataNaRm$steps,dataNaRm$date,mean,na.rm=T)
    StepsPerDayMean 
```

computing and reporting the **new** median number of **steps per day**:
```{r}
    StepsPerDayMedian <- tapply(dataNaRm$steps,dataNaRm$date,median,na.rm=T)
    StepsPerDayMedian
```

calculating the **new** mean number of **total steps over all days**:
```{r}
    MEAN <- mean(tapply(dataNaRm$steps,dataNaRm$date,sum,na.rm=T))
    MEAN
```

calculating the **new** median number of **total steps over all days**:
```{r}
    MEDIAN <- median(tapply(dataNaRm$steps,dataNaRm$date,sum,na.rm=T))
    MEDIAN
```

Replacing missing values with mean interval values centralizes the distribution. In the earlier analysis (no imputing) a day contained either NA observations only or none. Hence these days were not included ealier but now they are and they all have the value 10766.19 because it is the sum of all interval averages. This results in mean and media of **`r MEAN`** and **`r MEDIAN`**. 


## Are there differences in activity patterns between weekdays and weekends?


* transforming date data to date variable:
```{r}
    dataNaRm <- transform(dataNaRm,date=as.Date(strptime(date,"%Y-%m-%d")))
```

* adding weekday column:
```{r}
    dataNaRm$wkday <- weekdays(dataNaRm$date)
```

* *optional replacing german with english weekday names:*
```{r}
    dataNaRm$wkday[dataNaRm$wkday=="Montag"] <- "Monday"
    dataNaRm$wkday[dataNaRm$wkday=="Dienstag"] <- "Tuesday"
    dataNaRm$wkday[dataNaRm$wkday=="Mittwoch"] <- "Wednesday"
    dataNaRm$wkday[dataNaRm$wkday=="Donnerstag"] <- "Thursday"
    dataNaRm$wkday[dataNaRm$wkday=="Freitag"] <- "Friday"
    dataNaRm$wkday[dataNaRm$wkday=="Samstag"]<-"Saturday"
    dataNaRm$wkday[dataNaRm$wkday=="Sonntag"]<-"Sunday"
```

* adding weekday/weekend column:
```{r}
    dataNaRm$wkend <- rep("weekday",nrow(dataNaRm))
    wkend <- dataNaRm$wkday=="Saturday"|dataNaRm$wkday=="Sunday"
    dataNaRm$wkend[wkend] <- "weekend"
```

* computing-interval means for weekday and weekend data:
```{r}
    WeekDayComparison <- aggregate(dataNaRm$steps ~ dataNaRm$interval + dataNaRm$wkend, 
                                   list(dataNaRm$interval), FUN=mean)
    colnames(WeekDayComparison) <- c("interval", "wkend", "steps")
```

* plotting the graph:
```{r}
    xyplot(WeekDayComparison$steps~WeekDayComparison$interval|WeekDayComparison$wkend,
    layout=c(1,2),xlab="Interval",ylab="Number of Steps",panel=function(x,y,..){
    panel.xyplot(x,y,type="l")}) 
```

Obviously during the week there is a high activitiy around 9 am when people are presumably on their way to work. The distribution for weekend activity in contrast in rather smooth in  comparison.